---
title: "UCNRS Wildfire Analysis"
output: html_document
date: "2025-07-17"
---

#Journal
Journal of Applied Ecology 

#Title
Wildfire disturbance and ecological cascades: teasing apart the direct and indirect effects of fire on tick populations

#Corresponding Author
Samantha Sambado (she/her) - sbsambado@ucsb.edu or ssambado@stanford.edu


##Code to recreate analysis & tables in main text


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(tidyverse)
library(piecewiseSEM)

## upload necessary data
# plot level data
field_data <- read.csv("data/fieldmeasurements_20250715.csv") %>% dplyr::select(-X)

# landscape level data
reserve_climate <- read.csv("data/landscape_prism_20250715.csv") %>% dplyr::select(-c(X, vpd_min_normal, vpd_max_normal))
reserve_distburn <- read.csv("data/landscape_disttofire_20250715.csv") %>% dplyr::select(-X)
reserve_percburned <- read.csv("data/landscape_percentburned_20250715.csv") 

# sem data
sem_data <- read.csv("data/sem_covariates_20250715.csv") %>% dplyr::select(-X) 


# wilcoxon data
ndvi <- read.csv("data/landscape_ndvi_20250715.csv") %>% dplyr::select(-X) 
nbr <- read.csv("data/landscape_nbr_20250715.csv") %>% dplyr::select(-X) 

```

#Analysis

##Summary - Ticks

```{r}
## total ticks
sum(field_data$tick_count)

## tick density by treatment
field_data %>% 
  group_by(treatment) %>% 
  summarise(tick_density_mean = round(mean(tick_density),2), 
            tick_density_se = round(sd(tick_density) / sqrt(n()),1)) 

## tick density by veg
field_data %>% 
  group_by(veg_category,treatment) %>% 
  filter(treatment == "Burn") %>%
  #filter(treatment != "Burn") %>% 
  summarise(tick_density_mean = round(mean(tick_density),2), 
            tick_density_se = round(sd(tick_density) / sqrt(n()),1)) 

## tick density by year
field_data %>% 
  group_by(year,treatment) %>% 
  filter(treatment == "Burn") %>%
  #filter(treatment != "Burn") %>% 
  summarise(tick_density_mean = round(mean(tick_density),2), 
            tick_density_se = round(sd(tick_density) / sqrt(n()),1)) 

```


##Wilcoxon rank sum - Ticks
```{r}
## between burn and unburned plots
ticks_burn <- field_data %>% filter(treatment == "Burn" & year == 2022) 

ticks_unburn <- field_data %>% filter(treatment == "Unburn"& year == 2022)

wilcox.test(ticks_burn$tick_density, ticks_unburn$tick_density)
# 2022 : W = 2432, p-value = 8.585e-05


## between vegetation types 
# grassland
ticks_burn_g <- field_data %>% filter(treatment == "Burn" & year == 2022 & veg_category == "Grassland") 

ticks_unburn_g <- field_data %>% filter(treatment == "Unburn"& year == 2022 & veg_category == "Grassland")

wilcox.test(ticks_burn_g$tick_density, ticks_unburn_g$tick_density)
# W = 380, p-value = 0.6997

# shrub
ticks_burn_s <- field_data %>% filter(treatment == "Burn" & year == 2022 & veg_category == "Scrub") 

ticks_unburn_s <- field_data %>% filter(treatment == "Unburn"& year == 2022 & veg_category == "Scrub")

wilcox.test(ticks_burn_s$tick_density, ticks_unburn_s$tick_density)
# W = 189.5, p-value = 0.02164

# forest
ticks_burn_f <- field_data %>% filter(treatment == "Burn" & year == 2022 & veg_category == "Forest") 

ticks_unburn_f <- field_data %>% filter(treatment == "Unburn"& year == 2022 & veg_category == "Forest")

wilcox.test(ticks_burn_f$tick_density, ticks_unburn_f$tick_density)
# W = 246, p-value = 0.0002037
```

##Wildfire characteristics at reserve level
```{r}
## % burned
reserve_percburned %>% 
  group_by(reserve) %>% 
  summarise(reserve_percentburn = round(reserve_percentburn,2))


## dist from plots to perimeter
reserve_distburn %>% 
  mutate(treatment = case_when(grepl("UNBURN", plot_name) ~ "Burn",
                               TRUE ~ "Unburn")) %>% 
  filter(treatment == "Burn") %>%
  #filter(treatment == "Unburn") %>% 
  group_by(reserve) %>% 
  summarise(dist_to_perim_mean = round(mean(distance_to_perim,na.rm = TRUE),2),
            dist_to_perim_se = round(sd(distance_to_perim) / sqrt(n()),1))
```

##SEM
(check this section)
Inclusive model 
```{r}
mod_inclusive <- psem(
  # local model
  lm(log(tick_density+1) ~ 
       soil_surface_severity_mean +
       woody_debris_mean + 
       reserve_percentburn + distance_to_perim +
       tmean_normal + ppt_mean_overall,
     data = sem_data),
  
  lmer(woody_debris_mean ~ 
         soil_surface_severity_mean +
         ppt_mean_overall + tmean_normal + 
         reserve_percentburn + distance_to_perim + (1|veg_category), data = sem_data),

  
  lm(soil_surface_severity_mean ~ 
       reserve_percentburn +  distance_to_perim  +
       ppt_mean_overall + tmean_normal, data = sem_data),
  
  lm(reserve_percentburn ~ 
       ppt_mean_overall + tmean_normal, data = sem_data)
)  

summary(mod_inclusive)
plot(mod_inclusive)

```

Plot-level model
```{r}
mod_plot <- psem(
  # local model
  lmer(log(tick_density+1) ~ 
       soil_surface_severity_mean +
       woody_debris_mean + bare_soil_char_mean + (1|reserve),
     data = sem_data),
  
  lmer(woody_debris_mean ~ 
         soil_surface_severity_mean + 
         (1|reserve), data = sem_data),

  
  lmer(bare_soil_char_mean ~ 
       soil_surface_severity_mean + 
         (1|reserve), data = sem_data))  

summary(mod_plot)
plot(mod_plot)
```

Landscape-level model
```{r}
mod_landscape <- psem(
  lmer(log(tick_density+1) ~ 
       reserve_percentburn + distance_to_perim +
       tmean_normal + ppt_mean_overall + (1|treatment),
     data = sem_data),
  
  lm(reserve_percentburn ~ 
       ppt_mean_overall + tmean_normal, data = sem_data)
)  

summary(mod_landscape)
plot(mod_landscape)
```

##Wilcoxon paired - Vegetation
```{r}
## only select burn data
ndvi_burn <- ndvi %>% filter(treatment == "Burn")


## ML
ndvi_burn_2020_ml <- ndvi_burn %>% filter(year == 2020 & reserve == "McLaughlin")
ndvi_burn_2021_ml <- ndvi_burn %>% filter(year == 2021 & reserve == "McLaughlin")
ndvi_burn_2022_ml <- ndvi_burn %>% filter(year == 2022 & reserve == "McLaughlin")

wilcox.test(ndvi_burn_2021_ml$NDVI, ndvi_burn_2020_ml$NDVI,  paired = TRUE)
# V = 19, p-value = 0.4316
# NEW V = 19, p-value = 0.4316

wilcox.test(ndvi_burn_2022_ml$NDVI, ndvi_burn_2020_ml$NDVI, paired = TRUE)
# V = 15, p-value = 0.2324
# NEW V = 32, p-value = 0.6953 (changed)

## QR
ndvi_burn_2020_qr <- ndvi_burn %>% filter(year == 2020 & reserve == "Quail Ridge")
ndvi_burn_2021_qr <- ndvi_burn %>% filter(year == 2021 & reserve == "Quail Ridge")
ndvi_burn_2022_qr <- ndvi_burn %>% filter(year == 2022 & reserve == "Quail Ridge")

wilcox.test(ndvi_burn_2021_qr$NDVI,ndvi_burn_2020_qr$NDVI, paired = TRUE)
# V = 0, p-value = 6.104e-05

wilcox.test(ndvi_burn_2022_qr$NDVI,ndvi_burn_2020_qr$NDVI,  paired = TRUE)
# V = 1, p-value = 0.0001221
# NEW V = 5, p-value = 0.0006104 (changed)

## HT
ndvi_burn_2020_ht <- ndvi_burn %>% filter(year == 2020 & reserve == "Hastings")
ndvi_burn_2021_ht <- ndvi_burn %>% filter(year == 2021 & reserve == "Hastings")
ndvi_burn_2022_ht <- ndvi_burn %>% filter(year == 2022 & reserve == "Hastings")

wilcox.test(ndvi_burn_2021_ht$NDVI, ndvi_burn_2020_ht$NDVI,  paired = TRUE)
# V = 20, p-value = 0.02155
# NEW V = 8, p-value = 0.001526 (changed)

wilcox.test(ndvi_burn_2022_ht$NDVI, ndvi_burn_2020_ht$NDVI, paired = TRUE)
# V = 34, p-value = 0.1514
# V = 1, p-value = 0.0001221 (changed)

## BC
ndvi_burn_2020_bc <- ndvi_burn %>% filter(year == 2020 & reserve == "Big Creek")
ndvi_burn_2021_bc <- ndvi_burn %>% filter(year == 2021 & reserve == "Big Creek")
ndvi_burn_2022_bc <- ndvi_burn %>% filter(year == 2022 & reserve == "Big Creek")

wilcox.test(ndvi_burn_2021_bc$NDVI,ndvi_burn_2020_bc$NDVI, paired = TRUE)
# V = 9, p-value = 0.002014
# NEW V = 0, p-value = 0.0007247 (changed)

wilcox.test(ndvi_burn_2022_bc$NDVI,ndvi_burn_2020_bc$NDVI,  paired = TRUE)
# V = 6, p-value = 0.0008545
# NEW V = 0, p-value = 0.0007247 (changed)
```

NBR
```{r}
## only select burn data
nbr_burn <- nbr %>% filter(treatment == "Burn")


## ML
nbr_ml <- nbr_burn %>% filter(reserve == "McLaughlin")
wilcox.test(nbr_ml$post_NBR, nbr_ml$pre_NBR,  paired = TRUE, conf.int = TRUE)
# V = 0, p-value = 0.001953

## QR
nbr_qr <- nbr_burn %>% filter(reserve == "Quail Ridge")
wilcox.test(nbr_qr$post_NBR, nbr_qr$pre_NBR,  paired = TRUE, conf.int = TRUE)
# V = 0, p-value = 6.104e-05

## HT
nbr_ht <- nbr_burn %>% filter(reserve == "Hastings")
wilcox.test(nbr_ht$post_NBR, nbr_ht$pre_NBR,  paired = TRUE, conf.int = TRUE)
# V = 0, p-value = 6.104e-05

## BC
nbr_bc <- nbr_burn %>% filter(reserve == "Big Creek")
wilcox.test(nbr_bc$post_NBR, nbr_bc$pre_NBR,  paired = TRUE, conf.int = TRUE)
# V = 0, p-value = 0.0006653


```
#Tables


##Table 1. 
```{r}

## climate

## % burned
reserve_percburned %>% 
  group_by(reserve) %>% 
  summarise(reserve_percentburn = round(reserve_percentburn,2))

## mean tick density and se
field_data %>% 
  group_by(reserve, treatment) %>% 
  filter(treatment != "Burn") %>% 
  #filter(treatment == "Burn") %>%
    summarise(tick_density_mean = round(mean(tick_density),2), 
            tick_density_se = round(sd(tick_density) / sqrt(n()),1)) 
```

##Table 2
```{r}
field_data %>% 
  group_by(reserve, veg_category, treatment) %>% 
  filter(treatment == "Burn") %>%
  #filter(treatment != "Burn") %>%
    summarise(tick_density_mean = round(mean(tick_density),1), 
            tick_density_se = round(sd(tick_density) / sqrt(n()),1)) 
```

##Table 3
```{r}
## only select burn data
ndvi_burn <- ndvi %>% filter(treatment == "Burn")


## ML
ndvi_burn_2020_ml <- ndvi_burn %>% filter(year == 2020 & reserve == "McLaughlin")
ndvi_burn_2021_ml <- ndvi_burn %>% filter(year == 2021 & reserve == "McLaughlin")
ndvi_burn_2022_ml <- ndvi_burn %>% filter(year == 2022 & reserve == "McLaughlin")


wilcox.test(ndvi_burn_2021_ml$NDVI, ndvi_burn_2020_ml$NDVI,  paired = TRUE)
# V = 19, p-value = 0.4316
# NEW
wilcox.test(ndvi_burn_2022_ml$NDVI, ndvi_burn_2020_ml$NDVI, paired = TRUE)
# V = 15, p-value = 0.2324
# NEW

## QR
ndvi_burn_2020_qr <- ndvi_burn %>% filter(year == 2020 & reserve == "Quail Ridge")
ndvi_burn_2021_qr <- ndvi_burn %>% filter(year == 2021 & reserve == "Quail Ridge")
ndvi_burn_2022_qr <- ndvi_burn %>% filter(year == 2022 & reserve == "Quail Ridge")

wilcox.test(ndvi_burn_2021_qr$NDVI,ndvi_burn_2020_qr$NDVI, paired = TRUE)
# V = 0, p-value = 6.104e-05
# NEW
wilcox.test(ndvi_burn_2022_qr$NDVI,ndvi_burn_2020_qr$NDVI,  paired = TRUE)
# V = 1, p-value = 0.0001221
# NEW

## HT
ndvi_burn_2020_ht <- ndvi_burn %>% filter(year == 2020 & reserve == "Hastings")
ndvi_burn_2021_ht <- ndvi_burn %>% filter(year == 2021 & reserve == "Hastings")
ndvi_burn_2022_ht <- ndvi_burn %>% filter(year == 2022 & reserve == "Hastings")

wilcox.test(ndvi_burn_2021_ht$NDVI, ndvi_burn_2020_ht$NDVI,  paired = TRUE)
# V = 20, p-value = 0.02155
# NEW

wilcox.test(ndvi_burn_2022_ht$NDVI, ndvi_burn_2020_ht$NDVI, paired = TRUE)
# V = 34, p-value = 0.1514
# NEW


## BC
ndvi_burn_2020_bc <- ndvi_burn %>% filter(year == 2020 & reserve == "Big Creek")
ndvi_burn_2021_bc <- ndvi_burn %>% filter(year == 2021 & reserve == "Big Creek")
ndvi_burn_2022_bc <- ndvi_burn %>% filter(year == 2022 & reserve == "Big Creek")

wilcox.test(ndvi_burn_2021_bc$NDVI,ndvi_burn_2020_bc$NDVI, paired = TRUE)
# V = 9, p-value = 0.002014
# NEW

wilcox.test(ndvi_burn_2022_bc$NDVI,ndvi_burn_2020_bc$NDVI,  paired = TRUE)
# V = 6, p-value = 0.0008545
# NEW


```
