---
title: "UCNRS Wildfire Figure"
output: html_document
date: "2025-07-16"
---
#Journal
Journal of Applied Ecology 

#Title
Wildfire disturbance and ecological cascades: teasing apart the direct and indirect effects of fire on tick populations

#Corresponding Author
Samantha Sambado (she/her) - sbsambado@ucsb.edu or ssambado@stanford.edu


##Code to recreate figures in main & supplemental text

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(readr)
library(tidyverse)
library(rnaturalearth)
library(sf)
library(ggrepel)
library(patchwork)
library(cowplot)
library(gridExtra)

## upload necessary data
# (some datafiles will be read in via a function and may not be listed here)

# shapefile of CA
california <- ne_states(returnclass = "sf") %>% filter(admin == "United States of America", name == "California")

## fire shp files
hen_fire_shp <- st_read("data/shapefiles/fireperimeter/hen_fire.shp", quiet = TRUE)
river_fire_shp <- st_read("data/shapefiles/fireperimeter/river_fire.shp", quiet = TRUE)
dolan_fire_shp <- st_read("data/shapefiles/fireperimeter/dolan_fire.shp", quiet = TRUE)

## dNBR
# this file is too large to put anywhere
dnbr_raster <- read.csv("data/landscape_dnbrraster_20250715.csv") %>% 
  dplyr::select(-X) %>% 
  mutate(dNBR_class = factor(dNBR_class,
                       levels = c("Unburned",
                                  "Low Severity",
                                  "Moderate-Low Severity",
                                  "Moderate-High Severity",
                                  "High Severity")))

## field data
field_data <- read.csv("data/fieldmeasurements_20250715.csv") %>% dplyr::select(-X)

## ndvi
ndvi <- read.csv("data/landscape_ndvi_20250715.csv") %>% dplyr::select(-X) 

## reserve shp files
ml_shp <- st_read("data/shapefiles/reserves/mclaughlin/McLaughlin_Boundary.shp") %>% st_transform(crs = 4326)
qr_shp <- st_read("data/shapefiles/reserves/quailridge/Quail_Ridge_Boundary.shp") %>% st_transform(crs = 4326)
ht_shp <- st_read("data/shapefiles/reserves/hastings/Hastings_Boundary.shp") %>% st_transform(crs = 4326)
bc_shp <- st_read("data/shapefiles/reserves/bigcreek/Landels_Hill_Big_Creek_Boundary.shp") %>% st_transform(crs = 4326)
```

#Main Text

##Fig 1

Panel A
```{r}
## step 1. define box around study sites
bbox_poly <- st_as_sfc(st_bbox(c(xmin = -123, xmax = -121, ymin = 36, ymax = 39), 
                               crs = st_crs(california)))
# this helps aesthetics with line thickness
bbox_line <- st_boundary(bbox_poly)

## step 2. map inset of CA 
ca_inset <- ggplot() +
  geom_sf(data = california, fill = "grey90", color = "black") +
  geom_sf(data = bbox_line, fill = NA, color = "#f03b20") +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))


## step 3. make map of study sites
# need to alter lat/lon nudge for final figure
plots_map <- plots_sf %>% 
  group_by(reserve) %>% 
  slice(1) %>% 
  mutate(label_x = case_when(reserve == "Big Creek" ~ st_coordinates(geometry)[,1] - 0.03,
                             reserve == "McLaughlin" ~ st_coordinates(geometry)[,1] + 0.283, 
                             TRUE ~ st_coordinates(geometry)[,1]),
         label_y = case_when(reserve == "McLaughlin" ~ st_coordinates(geometry)[,2] + .0,
                             TRUE ~ st_coordinates(geometry)[,2]))


fire_perim_map <- ggplot() +
  geom_sf(data = california, fill = "grey90", color = "black") +
  geom_sf(data = hen_fire_shp, fill = "#f03b20", color = "black") +
  geom_sf(data = river_fire_shp, fill = "#f03b20", color = "black") + 
  geom_sf(data = dolan_fire_shp, fill = "#f03b20", color = "black") +
  geom_label_repel(data = plots_map, 
                   aes(x = label_x,y = label_y, label = reserve),
                   box.padding = 3,size = 6, fontface = "bold") +
  geom_sf(data = plots_clean_sf, shape = 24, fill = "#ffeda0", color = "black",size = 4) +
  coord_sf(xlim = c(-123, -121), ylim = c(36, 39)) +
  theme_light() +
  annotation_scale(location = "bl", width_hint = .2) +
  annotation_north_arrow(location = "bl", which_north = "true",
                         pad_x = unit(.01, "in"), pad_y = unit(.3, "in"),
                         style = north_arrow_fancy_orienteering) +
  labs(x = "Longitude", y = "Latitude") +
  theme(axis.title = element_text(face = "bold", size = 14),
        axis.text = element_text(size = 12))
  

```

Panel B
```{r}
## step 1. make individual plots

## McLaughlin
dnbr_plot_ml <- dnbr_raster  %>% 
  filter(reserve == "McLaughlin") %>% 
ggplot(aes(x = x, y = y, fill = dNBR_class)) +
 geom_raster()  +
  scale_fill_manual(values = c("#31a354" ,"#fff7bc", "#fdae61", "#f46d43","#d73027")) +
  labs(title = " McLaughlin", 
       fill = "Fire Severity") +
  theme_void() +  
  theme(title = element_text(face = "bold", size = 13),
        legend.position = c(.25,.25),#"bottom",.15
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.6, "cm"),
        plot.background = element_rect(color = "black", size = 1),
        axis.text = element_blank(), 
        axis.title = element_blank()) +
  coord_fixed(ratio = 1)

## Quail Ridge
dnbr_plot_qr <- dnbr_raster  %>% 
  filter(reserve == "Quail Ridge") %>% 
ggplot(aes(x = x, y = y, fill = dNBR_class)) +
 geom_raster()  +
  scale_fill_manual(values = c("#31a354" ,"#fff7bc", "#fdae61", "#f46d43","#d73027")) +
  labs(title = " Quail Ridge", 
       fill = "Fire Severity") +
  theme_void() +  
  theme(title = element_text(face = "bold", size = 13),
        legend.position = c(.25,.25),#"bottom",.15
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.6, "cm"),
        plot.background = element_rect(color = "black", size = 1),
        axis.text = element_blank(), 
        axis.title = element_blank()) +
  guides(fill = FALSE) +
  coord_fixed(ratio = 1)

## Hastings
dnbr_plot_ht <- dnbr_raster  %>% 
  filter(reserve == "Hastings") %>% 
ggplot(aes(x = x, y = y, fill = dNBR_class)) +
 geom_raster()  +
  scale_fill_manual(values = c("#31a354" ,"#fff7bc", "#fdae61", "#f46d43","#d73027")) +
  labs(title = " Hastings", 
       fill = "Fire Severity") +
  theme_void() +  
  theme(title = element_text(face = "bold", size = 13),
        legend.position = c(.25,.25),
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.6, "cm"),
        plot.background = element_rect(color = "black", size = 1),
        axis.text = element_blank(), 
        axis.title = element_blank()) +
  guides(fill = FALSE) +
  coord_fixed(ratio = 1)

## Big Creek
dnbr_plot_bc <- dnbr_raster  %>% 
  filter(reserve == "Big Creek") %>% 
ggplot(aes(x = x, y = y, fill = dNBR_class)) +
 geom_raster()  +
  scale_fill_manual(values = c("#31a354" ,"#fff7bc", "#fdae61", "#f46d43","#d73027")) +
  labs(title = " Big Creek", 
       fill = "Fire Severity") +
  theme_void() +  
  theme(title = element_text(face = "bold", size = 13),
        legend.position = c(.25,.25),
        legend.text = element_text(size = 10), 
        legend.key.size = unit(0.6, "cm"),
        plot.background = element_rect(color = "black", size = 1),
        axis.text = element_blank(), 
        axis.title = element_blank()) +
  guides(fill = FALSE) +
  coord_fixed(ratio = 1)
```

Combine panels
```{r}
# get rid of margin
fire_perim <- fire_perim_map + theme(plot.margin = margin(0, 0, 0, 0))

# create middle row
middle_row <- dnbr_plot_qr | dnbr_plot_ht

# create full reserve grid
reserves_grid <- dnbr_plot_ml / middle_row / dnbr_plot_bc +
  plot_layout(heights = c(1, 1, 1))

# combine both panels
both_panels <- fire_perim + reserves_grid + 
  plot_layout(widths = c(2.5, 2)) & 
  theme(plot.margin = margin(0, 0, 0, 0))

# format figure
both_panels_formatted <- cowplot::plot_grid(both_panels,
                                            ncol = 2,
                                            rel_widths = c(0.85, 0.15),
                                            align = "hv", axis = "tblr")

both_panels_formatted_annotated <- ggdraw() +
  draw_plot(both_panels_formatted) +
  draw_plot_label(label = c("A", "B"),
                  x = c(0.03, 0.45),
                  y = c(1,1),
                  fontface = "bold", size = 18) +
  draw_plot(ca_inset, x = 0.07, y = 0.15, width = 0.2, height = 0.2) +
  # add connecting line for ML label 
  draw_line(x = c(.215, .255),y = c(.89, .892), color = "black", size = .5)

# make sure it saves
print(both_panels_formatted_annotated)

## now officially save
ggsave(file = "figures/fig_1.png",
       plot = both_panels_formatted_annotated,
       width = 14, height = 10, dpi = 600)
```


##Fig 2

Panel A
```{r}
## format data
# summarise by veg category
fig2a_summary <- field_data %>% 
  group_by(treatment, veg_category) %>% 
  summarise(tick_density_mean = mean(tick_density, na.rm = TRUE),
            tick_density_se = sd(tick_density) / sqrt(n())) %>% 
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest")))

## plot it
fig2a <- fig2a_summary %>% 
  ggplot(aes(x = veg_category, y = tick_density_mean, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, color = "black") +
  geom_errorbar(aes(ymin = tick_density_mean - tick_density_se,
                    ymax = tick_density_mean + tick_density_se),
                position = position_dodge(width = .7), width = .2) +
  labs(x = NULL, y = expression(bold("Mean Tick Density per 100" ~ m^2)),
                                fill = "Treatment", tag = "A") +
         theme_classic(base_size = 16) +
  scale_fill_manual(values = c("#f03b20", "#91bfdb")) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 16),
        plot.tag = element_text(face = "bold", size = 18),
        axis.text.x = element_text(face = "bold")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), limits = c(0, NA)) +
  geom_text(aes(x = 1, y = 4.1, label = "p = 0.70"), 
            position = position_nudge(x = 0), size = 5, fontface = "bold", color = "black") +
  geom_text(aes(x = 2, y = 7.05, label = "p = 0.02*"), 
            position = position_nudge(x = 0), size = 5, fontface = "bold", color = "black") +
  geom_text(aes(x = 3, y = 6.3, label = "p < 0.001***"), 
            position = position_nudge(x = 0), size = 5, fontface = "bold", color = "black")
  
```


Panel B
```{r}
## format data
# summarise by veg category
fig2b_summary <- ndvi  %>% 
  filter(treatment == "Burn") %>% 
  group_by(veg_category) %>% 
  summarise(ndvi_basemedian = median(NDVI, na.rm = TRUE)) %>% 
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest"))) 
## plot it

fig2b <- ndvi %>% 
  filter(treatment == "Burn") %>% 
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest")),
         reserve = factor(reserve,
                          labels = c("BC", "HT", "ML", "QR"))) %>% 
  ggplot(aes(x = as.factor(year), y = NDVI)) +
  geom_hline(data = fig2b_summary, aes(yintercept = ndvi_basemedian), linetype = "dashed") +
  geom_boxplot(width = .5, fill = "grey85") + 
  geom_jitter(aes(color = reserve), width = .07, size = 2) +
  scale_color_viridis_d(direction = -1) +
  facet_wrap(~fct_relevel(veg_category, "Forest", "Shrubland", "Grassland"), ncol = 1) +
  labs(x = NULL, y = "NDVI", color = "Reserve", tag = "B") +
  theme_classic(base_size = 16) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.title = element_text(face = "bold", size = 16),
        plot.tag = element_text(face = "bold", size = 18),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  guides(color = guide_legend(override.aes = list(size = 3)))

```


Combine
```{r}
fig_2 <- ggarrange(fig2a, fig2b,
                   ncol = 2, align = "hv", 
                   widths = c(.40, .60))

ggsave(fig_2, file = "figures/fig_2.png", width = 11, height = 6, dpi = 600)
```


##Fig 3 
Need to add.


##Supplemental

##Fig S1
```{r}
## step 1. format data 
plots_spatial <- read.csv("data/plotscoords_spatial_20250715.csv") %>% 
  dplyr::select(-X) %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  mutate(plot_name = str_trim(plot_name))

ml_plots <- plots_spatial %>% filter(reserve == "McLaughlin")
ht_plots <- plots_spatial %>% filter(reserve == "Hastings")
bc_plots <- plots_spatial %>% filter(reserve == "Big Creek")

# QR bounday shp has a lot of internal lines that look messy
qr_kml <- st_read("data/shapefiles/reserves/quailridge_boundary2014_nointernallines.kml")
# get rid of points
quail_poly <- qr_kml[qr_kml$Name == "Quail Ridge Reserve", ]


## step 2. write function to give plots points padding
expand_bbox <- function(bbox_num, padding_ratio = 0.01) {
    xmin <- bbox_num[1]
  ymin <- bbox_num[2]
  xmax <- bbox_num[3]
  ymax <- bbox_num[4]
  
  xpad <- (xmax - xmin) * padding_ratio
  ypad <- (ymax - ymin) * padding_ratio
  
  
  c(xmin - xpad,
    ymin - ypad,
    xmax + xpad,
    ymax + ypad)
}

## step 3. write function to produce consistent plots
make_reserve_map <- function(reserve_name, reserve_shp, plots_sf) {
  reserve_bbox <- st_bbox(reserve_shp)
  reserve_bbox_num <- as.numeric(reserve_bbox)
  padded_bbox <- expand_bbox(reserve_bbox_num, 0.01)
  
  ggplot() +
    geom_sf(data = ca, fill = "grey99", color = NA) +
    geom_sf(data = reserve_shp, fill = "#d8e2dc", color = "black", size = 0.3) +
    geom_sf(data = plots_sf %>% filter(reserve == reserve_name),
            aes(color = treatment, shape = veg_type),
            size = 2) +
    coord_sf(xlim = c(padded_bbox[1], padded_bbox[3]),
             ylim = c(padded_bbox[2], padded_bbox[4]),
      expand = FALSE) +
    scale_color_manual(values = c("#E65100", "#1E88E5")) +
    scale_shape_manual(values = c("Grassland" = 16, 
                                  "Chamise" = 15,
                                  "Scrub" = 15,
                                  "Redwood" = 17,
                                  "Oak Woodland" = 17)) +
    labs(title = reserve_name, color = "Treatment", shape = "Habitat") +
    annotation_scale(location = "bl", width_hint = 0.2) +
    annotation_north_arrow(location = "bl", which_north = "true",
                           pad_x = unit(0.01, "in"), pad_y = unit(0.3, "in"),
                           style = north_arrow_fancy_orienteering) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.title = element_text(face = "bold"),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.background = element_rect(color = "black", size = 1),
      panel.border = element_rect(color = "black", fill = NA)
    )
}

## step 4. apply function 
ml_map <- make_reserve_map("McLaughlin", ml_shp, ml_plots)
qr_map <- make_reserve_map("Quail Ridge", qr_poly, qr_plots)
ht_map <- make_reserve_map("Hastings", ht_shp, ht_plots)
bc_map <- make_reserve_map("Big Creek", bc_shp, bc_plots)


## step 5. arrange plots
fig_s1 <- (ml_map + qr_map) / (ht_map + bc_map) +
  plot_layout(widths = c(1, 1), heights = c(0.8, 0.8))

ggsave(file = "figures/fig_s1.png", fig_s1,
       width = 14, height = 10, dpi = 300)
```

##Fig S2

Panel A
```{r}
figS2a_summary <- field_data %>% 
  group_by(treatment, veg_category, year) %>% 
  summarise(tick_density_mean = mean(tick_density), 
            tick_density_se = sd(tick_density) / sqrt(n())) %>% 
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest")))

figS2a <- figS2a_summary %>% 
  mutate(reserve = as.factor(reserve,
                          levels = c("McLaughlin", "Quail Ridge", "Hastings", "Big Creek"))) %>% 
  ggplot(aes(x = veg_category, y = tick_density_mean, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, color = "black") +
  geom_errorbar(aes(ymin = tick_density_mean - tick_density_se,
                    ymax = tick_density_mean + tick_density_se),
                position = position_dodge(width = 0.7), width = 0.2) +
  labs(x = NULL, y = expression(bold("Mean Tick Density per 100" ~ m^2)),
                                fill = "Treatment", tag = "A") +
         theme_classic(base_size = 16) +
  scale_fill_manual(values = c("#f03b20", "#91bfdb")) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(face = "bold"),
        plot.tag = element_text(face = "bold", size = 18),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), limits = c(0, NA)) +
  facet_wrap(~year, scales = "free_y")
```

Panel B
```{r}
figS2b_summary <- field_data %>% 
  group_by(treatment, veg_category, reserve) %>% 
  summarise(tick_density_mean = mean(tick_density), 
            tick_density_se = sd(tick_density) / sqrt(n())) %>% 
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest")))


figS2b <- figS2b_summary %>% 
  mutate(reserve = factor(reserve,
                          levels = c("McLaughlin", "Quail Ridge", "Hastings", "Big Creek"))) %>% 
  ggplot(aes(x = veg_category, y = tick_density_mean, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7, color = "black") +
  geom_errorbar(aes(ymin = tick_density_mean - tick_density_se,
                    ymax = tick_density_mean + tick_density_se),
                position = position_dodge(width = 0.7), width = 0.2) +
  labs(x = NULL, y = expression(bold("Mean Tick Density per 100" ~ m^2)),
                                fill = "Treatment", tag = "B") +
         theme_classic(base_size = 16) +
  scale_fill_manual(values = c("#f03b20", "#91bfdb")) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(face = "bold"),
        plot.tag = element_text(face = "bold", size = 18),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), limits = c(0, NA)) +
  facet_wrap(~reserve, scales = "free_y")
```


Combine
```{r}
figS2 <- ggarrange(figS2a +
            theme(legend.background = element_blank(),
                  legend.box.background = element_blank(), 
                  legend.key = element_blank()),
          figS2b +
            theme(legend.background = element_blank(),
                  legend.box.background = element_blank(), 
                  legend.key = element_blank()), 
          ncol = 1,
          common.legend = TRUE, legend = "bottom",
          heights = c(.40, .60)) + theme(plot.margin = margin(0, 0, 0, 0))


ggsave(file = "figures/fig_s2.png", figS2,
       width = 14, height = 12, dpi = 300,
       bg = "transparent")
```


##Fig S3
```{r}
## step 1. upload data
distance <- read.csv("data/landscape_disttofire_20250715.csv") %>% dplyr::select(-X) %>% 
  mutate(treatment = case_when(grepl("UNBURN", plot_name) ~ "Burn",
         TRUE ~ "Unburn"))

## step 2. plot
figS3 <- distance %>% 
  mutate(reserve = factor(reserve,
                          levels = c("Big Creek", "Hastings", "McLaughlin", "Quail Ridge"),
                          labels = c("BC", "HT", "ML", "QR"))) %>% 
  ggplot(aes(x = reserve, y = distance_to_perim), fill = reserve) +
  geom_boxplot(aes(fill = reserve)) +
  geom_jitter(width = .1, alpha = .5) +
  scale_fill_viridis_d(direction = -1) +
  labs(x = NULL, y = "Distance from plot\n to fire perimeter (m)") +
  theme_bw()+
  theme(axis.title = element_text(face = "bold", size = 14),
        axis.text = element_text(face = "bold", size = 12),
        plot.margin = margin(5, 5, 5, 5),
        strip.text = element_text(face = "bold", size = 14)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)), limits = c(0, NA)) +
  facet_wrap(~treatment) +
  guides(fill = FALSE)

ggsave(figS3, file = "figures/fig_s3.png", width = 7, height = 5, dpi = 600)
```

##Fig S4

Panel A
```{r}
## step 1. upload data
prism <- read.csv("data/landscape_prism_20250715.csv") %>% dplyr::select(-X)

## step 2. plot
figS4a <- prism %>% 
  mutate(reserve = factor(reserve,
                          levels = c("Big Creek", "Hastings", "McLaughlin", "Quail Ridge"),
                          labels = c("BC", "HT", "ML", "QR"))) %>% 
  ggplot(aes(x = reserve, y = tmean_normal)) +
  geom_jitter(alpha = .3, width = .2) +
  geom_boxplot(aes(fill = reserve)) +
  scale_fill_viridis_d(direction = -1) +
  labs(x = NULL, y = "Mean Temperature (°C)", tag = "A") +
  theme_bw(base_size = 16) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        plot.tag = element_text(face = "bold", size = 18),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  guides(fill = FALSE)
  
```

Panel B
```{r}
figS4b <- prism %>% 
  mutate(reserve = factor(reserve,
                          levels = c("Big Creek", "Hastings", "McLaughlin", "Quail Ridge"),
                          labels = c("BC", "HT", "ML", "QR"))) %>% 
  ggplot(aes(x = reserve, y = vpd_mean_normal)) +
  geom_jitter(alpha = .3, width = .2) +
  geom_boxplot(aes(fill = reserve)) +
  scale_fill_viridis_d(direction = -1) +
  labs(x = NULL, y = "Mean VPD (kPA)", tag = "B") +
  theme_bw(base_size = 16) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        plot.tag = element_text(face = "bold", size = 18),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  guides(fill = FALSE)
```

Panel C
```{r}
figS4c <- prism %>% 
  mutate(reserve = factor(reserve,
                          levels = c("Big Creek", "Hastings", "McLaughlin", "Quail Ridge"),
                          labels = c("BC", "HT", "ML", "QR"))) %>% 
  pivot_longer(cols = c("ppt_meanpost.firewinter_1":"ppt_meanpre.firewinter"),
               names_to = "season",
               values_to = "ppt_mm") %>% 
  mutate(reserve = factor(reserve,
                          labels = c("BC", "HT", "ML", "QR")),
         season = factor(season,
                         levels = c("ppt_meanpre.firewinter","ppt_meanpost.firewinter_1", "ppt_meanpost.firewinter_2"),
                         labels = c("2020", "2021", "2022"))) %>% 
  ggplot(aes(x = reserve, y = ppt_mm)) +
  geom_jitter(alpha = .3, width = .2) +
  geom_boxplot(aes(fill = reserve)) +
  scale_fill_viridis_d(direction = -1) +
  labs(x = NULL, y = "Winter Cum. Precip. (mm)", tag = "C") +
  theme_bw(base_size = 16) +
  theme(legend.position = "bottom",
        legend.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        plot.tag = element_text(face = "bold", size = 18),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  guides(fill = FALSE) +
  facet_wrap(~season)

```

Combine them
```{r}
## keep separate for now because it's easier to arrange
figs4ab <- ggarrange(figS4a, figS4b, align = "hv")
ggsave(figs4ab, file = "figures/fig_s4ab.png", width = 7, height = 4, dpi = 600)

ggsave(figS4c, file = "figures/fig_s4c.png", width = 7, height = 4, dpi = 600)
```

##Fig S5

```{r}
## step 1. upload data
ndvi <- read.csv("data/landscape_ndvi_20250715.csv") %>% dplyr::select(-X)

## step 2.format data
figS5_summary <- ndvi %>% 
  filter(treatment == "Burn" & year == 2020) %>% 
  group_by(reserve, veg_category) %>% 
  summarise(ndvi_basemedian = median(NDVI, na.rm = TRUE)) %>% 
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest"))) 


## step 3. plot it
figS5 <- ndvi %>% 
  filter(treatment == "Burn") %>%   
  mutate(veg_category = factor(veg_category,
                               levels = c("Grassland",
                                          "Scrub",
                                          "Forest"),
                               labels = c("Grassland",
                                          "Shrubland",
                                          "Forest"))) %>% 
  ggplot(aes(x = factor(year), y = NDVI)) +
  geom_hline(data = fig_s4_summary, aes(yintercept = ndvi_basemedian),
             color = "#f03b20", linetype = "dashed") +
  geom_boxplot(width = .5, aes(fill = reserve)) +
  scale_fill_viridis_d(direction = -1) +
  labs(x = NULL, y = "NDVI") +
  theme_bw(base_size = 16) +
  theme(
        axis.text.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        strip.text = element_text(face = "bold"),
        strip.background = element_rect(fill = "grey95")) +
  guides(fill = FALSE) +
  facet_grid(veg_category ~ reserve)

ggsave(figS5, file = "figures/fig_s5.png", width = 7, height = 5, dpi = 600)
```